On the value of genomic wastewater surveillance information for pandemic response

### 2023-05-31 paper/manuscript.qmd
updates manuscript

---
title: Modeling costs and benefits of genomic wastewater surveillance
# subtitle: Supplementary appendix draft
#author:
#  - name: Pedro Nascimento de Lima
#    email: plima@rand.org
#    affiliations: 
#        - id: rand
#          name: RAND Corporation
#          department: Engineering and Applied Sciences
#          address: 1776 Main Street
#          city: Santa Monica
#          state: CA
#          postal-code: 90401
#    attributes:
#        corresponding: true
#  - name: and colleagues
#    email: others@rand.org
#    affiliations:
#        - ref: rand
abstract: |
  Introduces a model to assess the value provided by wastewater genomic surveillance network. This document is meant to be used as a supplementary appendix of the paper.
keywords: 
  - Genomic Surveillance
  - Pandemic Response
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  elsevier-html:
    keep-tex: true
    mermaid:
      theme: neutral
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  html:
    keep-tex: true
    mermaid:
      theme: neutral
---

@@

This paper develops a stochastic metapopulation disease transmission model and uses it to assess the conditions under which WBS systems can provide value to local public health decision makers who respond to a pandemic using surveillance data. We evaluate a range of scenarios that represent different potential pathogens and highlight scenarios under which WBS systems may or may not be cost-effective.
<!--
Our results highlight that the value of WBS systems is contingent on a set of biological and societal parameters, including pathogen characteristics (i.e., transmissibility and severity), as well as the geographic scale of 
-->

@@

We start with a standard disease transmission model describing person-to-person transmission of an infectious pathogen across $n$ jurisdictions. The state-space of the model is comprised of a set of health statuses $\mathbf{S}, \mathbf{E}, \mathbf{P}, \mathbf{I}, \mathbf{R}$, each a $1 \times n$ vector representing the population within each compartment, respectively representing the number of Susceptibles, Exposed, Pre-Symptomatic, Infected and Removed[^1]. @eq-ode1 describes a system of differential equations representing the model, and @tbl-parameters contains parameter definitions, values and their sources.

@@

$$ 
\mathbf{L}_t^* = \min{(cI_{t-\Delta}, l_{max})}
$$ {#eq-targetnpi}
where $\Delta$ is the delay imposed by the surveillance method used to monitor the disease. This formulation approximates phased mitigation plans adopted during the COVID-19 pandemic, where policymakers introduce or remove interventions depending on disease prevalence or other severity indicators. While public health decision-makers strive to use up-to-date information to introduce and remove NPIs, they do not have perfect and instantaneous information on disease prevalence; hence actual policy behaviors lag epidemiological indicators. In continuous time, we represent this process with a stock vector $\mathbf{L}$ representing the NPI level in effect

@@

where $a$ represents the adjustment time for NPIs for decision making reasons, whereas $\Delta$ represents the lag imposed by surveillance methods.

>>>>

---
title: On the value of genomic wastewater surveillance information for pandemic response
subtitle: A model-based analysis
#author:
#  - name: Pedro Nascimento de Lima
#    email: plima@rand.org
#    affiliations: 
#        - id: rand
#          name: RAND Corporation
#          department: Engineering and Applied Sciences
#          address: 1776 Main Street
#          city: Santa Monica
#          state: CA
#          postal-code: 90401
#    attributes:
#        corresponding: true
#  - name: and colleagues
#    email: others@rand.org
#    affiliations:
#        - ref: rand
abstract: |
  While wastewater-based genomic surveillance (WBS) systems may provide an early warning system for pandemic response, its adoption is still incomplete and it is unclear whether public health departments should allocate a larger budget to those systems. This paper introduces a model to assess the value provided by wastewater genomic surveillance network and clarifies the conditions under which WBS systems provide the most value to public health decision makers.
keywords: 
  - Genomic Surveillance
  - Pandemic Response
# date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  elsevier-html:
    keep-tex: true
    mermaid:
      theme: neutral
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  html:
    keep-tex: true
    mermaid:
      theme: neutral
---

@@

This paper develops a stochastic metapopulation disease transmission model and uses it to assess the conditions under which WBS systems can provide value to local public health decision makers who respond to a pandemic using surveillance data. In the model, WBS systems help decision-makers by reducing the lag between the latent disease prevalence and the epidemiological time-series (i.e., case counts) decision-makers often use to introduce interventions targeted to reduce disease transmission. We evaluate a range of scenarios that represent different potential pathogens and highlight scenarios under which WBS systems may or may not be cost-effective.
<!--
Our results highlight that the value of WBS systems is contingent on a set of biological and societal parameters, including pathogen characteristics (i.e., transmissibility and severity), as well as the geographic scale of 
-->

@@

Our analysis starts with a standard disease transmission model describing person-to-person transmission of an infectious pathogen across $n$ jurisdictions. The state-space of the model is comprised of a set of health statuses $\mathbf{S}, \mathbf{E}, \mathbf{P}, \mathbf{I}, \mathbf{R}$, each a $1 \times n$ vector representing the population within each compartment, respectively representing the number of Susceptibles, Exposed, Pre-Symptomatic, Infected and Removed[^1]. @eq-ode1 describes a system of differential equations representing the model, and @tbl-parameters contains parameter definitions, values and their sources.

@@

$$ 
\mathbf{L}_t^* = \min{(cI_{t-s_{lag}}, l_{max})}
$$ {#eq-targetnpi}
where $s_{lag}$ is a surveillance lag imposed by the surveillance method used to monitor the disease. This formulation approximates phased mitigation plans adopted during the COVID-19 pandemic, where policymakers introduce or remove interventions depending on disease prevalence or other severity indicators. While public health decision-makers strive to use up-to-date information to introduce and remove NPIs, they do not have perfect and instantaneous information on disease prevalence; hence actual policy behaviors lag epidemiological indicators. In continuous time, we represent this process with a stock vector $\mathbf{L}$ representing the NPI level in effect

@@

where $a$ represents the adjustment time for NPIs for decision making and implementation reasons, whereas $s_{lag}$ represents the lag imposed by surveillance methods.
